Robust Winner Determination and Simultaneous Elicitation of Scoring Rules and Preferences

### 2019-02-20 draft.tex
Moved around and started alt def PMR

\newcommand{\profile}{\bm{v}}%(complete) profile
\newcommand{\pprofile}{\bm{p}}%partial profile
\newcommand{\w}{\bm{w}}%partial weights
\newcommand{\C}{\mathcal{C}}%partial profile
\newcommand{\powersetz}[1]{\mathscr{P}^*(#1)}

@@

\section{Social choice with partial information}\label{sec:background}
We now introduce some basic concepts.
We assume defined a set $A$ of $m$ alternatives (products, restaurants, movies, public projects, job candidates, etc.) and a set $\set{1, …, n}$ of agents (voters); each agent $j$ is associated to her “real” preference order $\pref_j$.
The real preferences of the agents are elements of $\linors$, the linear orders (connected, transitive, asymmetric relations) over the alternatives.
Following the social choice nomenclature, we call {\em profile} the association of a preference to each voter, denoted by $(\pref_1,\ldots,\pref_n)$, which we assume is fixed but unknown.
The profile is equivalently represented by $\profile=(v_1,\ldots,v_n)$ where $v_j(i) \in \set{1, \ldots, m}$ denotes the rank (position) of alternative $i$ in the preference order $\pref_j$. 
%With a little abuse of terminology, we will use the term profile to refer to either $\profile$ or to the preference relations, depending on the context.

Let $V$ be the set of possible preference profiles (the $n$-fold cartesian product of the linear orders over the alternatives).
A social choice function $f : V \rightarrow \powersetz{A}$ associates a profile with a set of winners, where $\powersetz{A}$ represents the set of subsets of $A$ except for the emptyset. (Sets are used for tied winners.)
Among the many possible social choice functions, we consider {\em positional scoring rules}, which attach weights to positions according to  the vector $(w_1, \ldots, w_m)$ (also called the scoring vector).
%A scoring rule associates each alternative to a score that is given by the sum of points obtained for each voter.
An alternative obtains a score that depends on the rank obtained in each of the preference orders \commentOC{Consider removing $v$ and $w$}:
\begin{align}
\label{eq:srule}
s(x; \profile, \w) = \sum_{j=1}^{n} w_{v_j(x)}
= \sum_{r=1}^{m} \alpha^{x}_r w_r 
\end{align}
where $\alpha^{x}_r$ is the number of times that alternative $x$ was ranked in the $r$-th position.
The winners are the alternatives with highest scores.
%We assume that the weights constitute a monotonic sequence: $w_{1} \geq w_{2} \geq \ldots \geq w_{m}$.

In this work we want to reason about partial preference information.
Our knowledge at a given time of the preference of voter $j$ is encoded by a partial order over the alternatives, thus a transitive and asymmetric binary relation, denoted by $\ppref_j$. 
In this work we assume that preference information is truthful, i.e. $a \ppref_j b ⇒ a \pref_j b$.
We use $\pinc$ to denote incomparability, that is $a \pinc_{j} b$ iff $a \nppref_j b \wedge b \nppref_j a$.
An incomplete profile $\pprofile = (\ppref_1, \ldots, \ppref_n)$ maps each voter to a partial preference.

A completion of $\ppref_j$ is any linear order $\pref$ that extends $\ppref_j$.
Let $C(\ppref_i) = \set{{\succ} \in \linors \suchthat {\ppref_j} \subseteq {\succ}}$ be the set of possible completions of $\ppref_i$.
We let $C(\pprofile)=C(\ppref_1)\times \ldots \times C(\ppref_n)$ be the set of complete profiles extending $p$.

\medskip
We also assume that the weights of the scoring rule are only partially specified.
Therefore the vector $(w_1,\ldots,w_m)$  is not known but we are given a set of constraints restraining the possible values that weights can take.
We consider a non-increasing sequence of weights:
\begin{align}
1=w_{1} \geq w_{2} \geq \ldots \geq w_{m}=0. \label{eq:monotone}
\end{align}
This is a natural assumption, as it is better to be ranked first than second, second than third, etc. 
Without loss of generality, we consider that $w_1=1$ and $w_m=0$.

@@

In general it can be difficult to set the weights in an appropriate way.
We assume that in addition of basic requirements (monotonicity and convexity), the chair (the person or the organization that is supervising the voting process) is able to specify additional preferences about how the social choice function should behave.
In this work we assume that preferences of the chair are encoded with linear constraints about the vector $\w$, relating the value of the weights of different positions.
We use $\C(\w^\text{p})$ to denote the set of weight vectors compatible with the preferences expressed by the chair about the scoring vector. \commentOC{Think about this.}

@@

The maximum regret is considered by assuming that an adversary both 1) can extend the partial profile $\pprofile$ into a complete profile, and 2) can instantiate the weights choosing among any weight vector in $C(\w^\text{p})$, where $\pprofile$ and $\w^\text{p}$ represent our knowledge so far.

We formalize the notion of minimax regret in multiple steps.
First of all, $\Regret(x, \profile, \w)$ is the loss or ``actual'' regret  of selecting $x$ as a winner instead of choosing the optimal alternative under $\profile$ and $\w$:
\[\Regret(x, \profile, \w) = \max_{y \in A} s(y; \profile,\w) - s(x; \profile, \w).\]

The pairwise max regret of $x$ relative to $y$ given partial profile $\pprofile$ and the space of weights $W$
$\PMR(x,y;\pprofile,W)$ is the worst-case loss under all possible realizations of the full profile {\em and} all possible instantiations of the weights:
\begin{align}
\PMR(x,y; \pprofile, W) & = \max_{\w \in W} \max_{\profile \in C(p)} s(y; \profile,\w) - s(x; \profile,\w).
\end{align}

Max regret $\MR(x;\pprofile,W)$ is the worst-case loss of $x$. It is the loss occurred by an adversarial selection of a complete profile $\profile$ extending $\pprofile$ and a selection of $\w \in W$ to maximize the loss between $x$ and the true winner under $\profile$ and $w$.
\begin{align}
\MR(x; \pprofile, W) & = \max_{y \in A} \PMR(x,y; \pprofile, W)\\
& = \max_{\w \in W} \max_{\profile \in C(p)} \Regret(x, \profile, \w).
\end{align}

@@

\subsubsection{Minimax regret without the convex assumption}
{\em (Still here? Or move it somewhere else?) }

Without the convex assumption, we cannot use $\hat{v}$ for the agents in $U^{-}$, but only for agents in $U^{+}$ and $U^{?}$.
%Let $\hat{v}_i$ be the linear order extending $\succ^{p}_i$ according to the above procedure.
Then PMR can be written as follows:
\begin{align*}
 &\PMR(x,y; \pprofile, W) =\\ 
 &\max_{\w \in W} \Big \{ \sum_{j \in U^-} [\max_{v_j \in C(\succ_j^p)} [w_{v_j(y)} \!-\! w_{v_j(x)}]] 
  \!+\!	 \sum_{j \in A^+ \cup  U^?} [w_{\hat{v}_j(y)} \!-\! w_{\hat{v}_j(x)}] \Big \} 
 \end{align*}
%Note that the second addendum inside the max do not depend on the choice of $\w$.
Consider the two addenda inside the ``max''. 
The first addendum is concerned with positioning of alternatives $x$ and $y$ for voters in $U^{-}$ for which we know that $x$ is preferred to $y$.
The second addendum is concerned with voters in $U^{+}$ and $U^{?}$.
We rewrite the second addendum as:
\begin{align*}
\sum_{j \in U^+ \cup  U^?} [w_{\hat{v}_j(y)} \!-\! w_{\hat{v}_j(x)}] 
= \sum_{i = 1}^{m} (\hat{\alpha}_{i}^{y} - \hat{\alpha}_{i}^{x}) w_{i}
\end{align*}
where $\hat{\alpha}_{i}^{x}$ is the number of times that $x$ is ranked  in position $i$  considering the profile $\hat{v}$ of voters in $U^+ \cup  U^?$.
%We compute pairwise maximum regret by considering binary variables $\{ B_{i} \}_{i=1,\ldots,m}$ to represent optimization choices related to where to position the alternatives.

\section{Interactive Elicitation} \label{sec:elicit}

We propose an incremental elicitation method based on minimax regret.
At each step, the system may either ask a question to one of the agents, or ask a question to the chair about the voting rule. 
The goal is to acquire relevant information to reduce minimax regret as quick as possible.
%Minimax regret can be used to guide the elicitation process.
As termination condition of elicitation, we can check whether minimax regret is lower than a threshold; if we wish optimality, we can perform elicitation until minimax regret drops to zero.

@@

>>>>

\newcommand{\profile}{\bm{v}}%(complete) profile
\newcommand{\pprofile}{{\bm{p}}}%partial profile
\newcommand{\w}{\bm{w}}
\newcommand{\pw}{W}%our knowledge about the weights
\newcommand{\C}{\mathcal{C}}%partial profile
\newcommand{\powersetz}[1]{\mathscr{P}^*(#1)}

@@

\section{Social choice with partial information}
\label{sec:background}
We now introduce some basic concepts.
We assume defined a set $A$ of $m$ alternatives (products, restaurants, movies, public projects, job candidates, etc.) and a set $\set{1, …, n}$ of agents (voters); each agent $j$ is associated to her “real” preference order $\pref_j$.
The real preferences of the agents are elements of $\linors$, the linear orders (connected, transitive, asymmetric relations) over the alternatives.
Following the social choice nomenclature, we call {\em profile} the association of a preference to each voter, denoted by $(\pref_1,\ldots,\pref_n)$.
The profile is equivalently represented by $\profile=(v_1,\ldots,v_n)$ where $v_j(i) \in \set{1, \ldots, m}$ denotes the rank (position) of alternative $i$ in the preference order $\pref_j$.

Let $V$ be the set of possible preference profiles (the $n$-fold cartesian product of the linear orders over the alternatives).
A social choice function $f : V \rightarrow \powersetz{A}$ associates a profile with a set of winners, where $\powersetz{A}$ represents the set of subsets of $A$ except for the emptyset. (Sets are used for tied winners.)
Among the many possible social choice functions, we consider {\em positional scoring rules}, which attach weights to positions according to  the vector $(w_1, \ldots, w_m)$ (also called the scoring vector).
%A scoring rule associates each alternative to a score that is given by the sum of points obtained for each voter.
An alternative obtains a score that depends on the rank obtained in each of the preference orders:
\begin{align}
\label{eq:srule}
s(x; \profile, \w) = \sum_{j=1}^{n} w_{v_j(x)}
= \sum_{r=1}^{m} \alpha^{x}_r w_r 
\end{align}
where $\alpha^{x}_r$ is the number of times that alternative $x$ was ranked in the $r$-th position.
The winners are the alternatives with highest scores.
%We assume that the weights constitute a monotonic sequence: $w_{1} \geq w_{2} \geq \ldots \geq w_{m}$.

In this work we assume fixed, but unknown, a profile $\profile^*$ (and corresponding voters preferences $\pref_j^*$) and a weight vector $\w^*$, and we want to reason about partial preference information concerning those objects.
Our knowledge at a given time of the preference of voter $j$ is encoded by a partial order over the alternatives, thus a transitive and asymmetric binary relation, denoted by $\ppref_j$. 
In this work we assume that preference information is truthful, i.e. $a \ppref_j b ⇒ a \pref_j^* b$.
We use $\pinc$ to denote incomparability, that is $a \pinc_{j} b$ iff $a \nppref_j b \wedge b \nppref_j a$.
An incomplete profile $\pprofile = (\ppref_1, \ldots, \ppref_n)$ maps each voter to a partial preference.

A completion of $\ppref_j$ is any linear order $\pref$ that extends $\ppref_j$.
Let $C(\ppref_j) = \set{{\succ} \in \linors \suchthat {\ppref_j} \subseteq {\succ}}$ be the set of possible completions of $\ppref_i$.
We let $C(\pprofile)=C(\ppref_1)\times … \times C(\ppref_n)$ be the set of complete profiles extending $p$. Note that $\profile^* \in C(\pprofile)$.

\medskip
We also assume that the weights of the scoring rule are only partially specified.
Therefore the vector $(w_1,\ldots,w_m)$  is not known but we are given a set of constraints restraining the possible values that weights can take.
We consider a decreasing sequence of weights:
\begin{align}
1=w_{1} > w_{2} > \ldots > w_{m}=0. \label{eq:monotone}
\end{align}
This is a natural assumption, as it is better to be ranked first than second, second than third, etc. 
Without loss of generality, we consider that $w_1=1$ and $w_m=0$. \commentOC{I suggest we say no more than this to justify this assumption. Otherwise, we might say something like the following. If the committee considers that, for example, “only the first ten positions matter” (so that $w_r = 0$ for $r > 10$), then we simply assume the committee is able to express this constraint at the start of the querying process, thus, that this information is known.}

@@

In general it can be difficult to set the weights in an appropriate way.
We assume that in addition of basic requirements (monotonicity and convexity), the chair (the person or the organization that is supervising the voting process) is able to specify additional preferences about how the social choice function should behave.
In this work we assume that preferences of the chair are encoded with linear constraints about the vector $\w$, relating the value of the weights of different positions.
We use $\pw \subseteq \mathcal{W}$ to denote the set of weight vectors compatible with the preferences expressed by the chair about the scoring vector.

@@

The maximum regret is considered by assuming that an adversary both 1) can extend the partial profile $\pprofile$ into a complete profile, and 2) can instantiate the weights choosing among any weight vector in $\pw$, where $\pprofile$ and $\pw$ represent our knowledge so far.

We formalize the notion of minimax regret in multiple steps.
First of all, denote by $s^*_{\profile, \w} = \max_{y \in A} s(y; \profile, \w)$ the score achieved by winning alternatives under $\profile$ and using weights $\w$. Now define $\Regret(x, \profile, \w)$ as the “regret” of selecting $x$ as a winner instead of choosing the optimal alternative under $\profile$ and $\w$:
\[\Regret(x, \profile, \w) = s^*_{\profile, \w} - s(x; \profile, \w).\]

%The pairwise max regret $\PMR(x,y;\pprofile,W)$ of $x$ relative to $y$ given partial profile $\pprofile$ and the set of weights $W$
%is the worst-case loss under all possible realizations of the full profile {\em and} all possible instantiations of the weights:
%\begin{align}
%\PMR(x,y; \pprofile, W) & = \max_{\w \in W} \max_{\profile \in C(\pprofile)} s(y; \profile,\w) - s(x; \profile,\w).
%\end{align}

Max regret $\MR(x;\pprofile,W)$ is the worst-case loss of $x$. It is the loss occurred by an adversarial selection of a complete profile $\profile$ extending $\pprofile$ and a selection of $\w \in W$ to maximize the loss between $x$ and the true winner under $\profile$ and $\w$.
\begin{align}
\MR(x; \pprofile, W) & = \max_{y \in A} \PMR(x,y; \pprofile, W)\\
& = \max_{\w \in W} \max_{\profile \in C(p)} \Regret(x, \profile, \w).
\end{align}

@@

\section{Interactive Elicitation} 
\label{sec:elicit}
We propose an incremental elicitation method based on minimax regret.
At each step, the system may either ask a question to one of the agents, or ask a question to the chair about the voting rule. 
The goal is to acquire relevant information to reduce minimax regret as quick as possible.
%Minimax regret can be used to guide the elicitation process.
As termination condition of elicitation, we can check whether minimax regret is lower than a threshold; if we wish optimality, we can perform elicitation until minimax regret drops to zero.

@@

\section{Minimax regret without the convex assumption}
Without the convex assumption, we cannot use $\hat{v}$ for the agents in $U^{-}$, but only for agents in $U^{+}$ and $U^{?}$.
%Let $\hat{v}_i$ be the linear order extending $\succ^{p}_i$ according to the above procedure.
Then PMR can be written as follows:
\begin{align*}
 &\PMR(x,y; \pprofile, W) =\\ 
 &\max_{\w \in W} \Big \{ \sum_{j \in U^-} [\max_{v_j \in C(\succ_j^p)} [w_{v_j(y)} \!-\! w_{v_j(x)}]] 
  \!+\!	 \sum_{j \in A^+ \cup  U^?} [w_{\hat{v}_j(y)} \!-\! w_{\hat{v}_j(x)}] \Big \} 
 \end{align*}
%Note that the second addendum inside the max do not depend on the choice of $\w$.
Consider the two addenda inside the ``max''. 
The first addendum is concerned with positioning of alternatives $x$ and $y$ for voters in $U^{-}$ for which we know that $x$ is preferred to $y$.
The second addendum is concerned with voters in $U^{+}$ and $U^{?}$.
We rewrite the second addendum as:
\begin{align*}
\sum_{j \in U^+ \cup  U^?} [w_{\hat{v}_j(y)} \!-\! w_{\hat{v}_j(x)}] 
= \sum_{i = 1}^{m} (\hat{\alpha}_{i}^{y} - \hat{\alpha}_{i}^{x}) w_{i}
\end{align*}
where $\hat{\alpha}_{i}^{x}$ is the number of times that $x$ is ranked  in position $i$  considering the profile $\hat{v}$ of voters in $U^+ \cup  U^?$.
%We compute pairwise maximum regret by considering binary variables $\{ B_{i} \}_{i=1,\ldots,m}$ to represent optimization choices related to where to position the alternatives.
